---
title: "Using Machine Learning to Predict Diabetes"
author: "Edward B. Quinn, PhD, MPH"
date: 07/30/2023
draft: false
categories: [R]
format: 
  html:
    toc: true
    toc-location: left
    toc-title: Contents
---

# Predicting Diabetes

One of the most exciting applications of machine learning is in the prediction of disease states. Here, I demonstrate the use of the newest modeling framework in R to compare many different types of models in the prediction of diabetes.

The data come from citation and link to full citation footnote. 

Let's load our libraries.

```{r}

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tidymodels))
suppressPackageStartupMessages(library(RCurl))
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(here))


tidymodels_prefer()

```


Read in the data and take a look at the first ten rows:

```{r}
#| message: false

df <- read_csv(here("r_projects/r_machine_learning/diabetes.csv"))

df

```


It is always a good idea to visualize the data as part of exploratory data analysis. 

```{r fig.width=20}

#| message: false
#| fig-width: 100
#| fig-height: 100
#| out-width: "1111%"


# ggpairs(df)

ggplot(df, aes(Age, BMI)) +
  geom_point()

```









